---
title: "Descriptive Statistics"
format: html
---


### Research Question and Variables to be Included

How does fasting influence insulin and cholesterol levels in adult men and women? 

In our final project we will look at gender, length of food fast (hours), length of coffee/tea fast (hours), length of alcohol fast(hours) to understand how different fasting aspect will affect two continuous outcome variables, insulin and cholesterol. 

We will also take into account individuals that have diabetes, as their insulin regulation is compromised. 

```{r}
library("here")
library("haven")
Demographics<-read_xpt(here("DEMO_L.xpt"))
Fasting<-read_xpt(here("FASTQX_L.xpt"))
Insulin<-read_xpt(here("INS_L.xpt"))
Cholesterol<-read_xpt(here("TCHOL_L.xpt"))
Diabetes<-read_xpt(here("DIQ_L.xpt"))


FullData_df<-Demographics|>
  full_join(Fasting, by = "SEQN")|>
  full_join(Insulin, by = "SEQN")|>
  full_join(Cholesterol, by = "SEQN")|>
  full_join(Diabetes, by = "SEQN")

Data_df<-FullData_df|> 
  select("SEQN", "RIAGENDR", "RIDAGEYR", "PHQ020", "PHACOFHR", "PHQ030", "PHAALCHR", "PHAFSTHR", "LBXTC", "LBXIN", "DIQ010", "DIQ050", "DIQ160")|>
  rename("Sequence" = SEQN,
         "Sex" = RIAGENDR,
         "Age" = RIDAGEYR,
        "Coffee_or_Tea_Factor" = PHQ020, "Hrs_CoffeeTea" = PHACOFHR, 
        "Alcohol_Factor" = PHQ030, "Hrs_Alcohol" = PHAALCHR,
        "Hrs_Food" = PHAFSTHR, 
        "Cholesterol" = LBXTC,
        "Insulin" = LBXIN,
        "Diabetes" = DIQ010, "Taking_Insulin" = DIQ050, "Prediabetes" = DIQ160)|>
  mutate(Sex = case_when(Sex == 1 ~"male", 
                         Sex == 2 ~ "female"),
         Coffee_or_Tea_Factor = case_when(Coffee_or_Tea_Factor == 1 ~ "yes",
                                          Coffee_or_Tea_Factor == 2 ~"no",
                                          .default = NA),
         Alcohol_Factor = case_when(Alcohol_Factor == 1 ~ "yes",
                                 Alcohol_Factor == 2 ~ "no", 
                                 .default = NA),
         Diabetes = ifelse(Diabetes == 1, "yes", "other"),
         Prediabetes = ifelse(Prediabetes == 1, "yes", "other"),
         Taking_Insulin = ifelse(Taking_Insulin == 1, "yes", "other"))

```

Cholesterol = mg/dL
Insulin = uU/mL

Coffee_or_Tea_Factor = Whether or not an individual fasted from coffee/tea
Same as alcohol 


### Single Variable Visualizations

Numerical and graphical descriptions for each variable 

### Bivariate Visualization 

Variables Jointly, bivariate between dependent and outcome 


### Summary of Trends Observed so far

trends observed after initial visualization 

```{r}
print(4)
```

